Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2014-11-28 manuscript.tex
Reordered and joined sentences for methods

\author{
    Leonardo Uieda\footnotemark[1]\footnotemark[2],
    Val\'eria C. F. Barbosa\footnotemark[2],
    Carla Braitenberg\footnotemark[3],
    and
    Vanderlei C. Oliveira Jr\footnotemark[2]
}

\address{
    \footnotemark[1]Universidade do Estado do Rio de Janeiro,
    Rio de Janeiro, Brazil.
    email: leouieda@gmail.com
    \\
    \footnotemark[2]Observat\'orio Nacional,
    Rio de Janeiro, Brazil.
    \\
    \footnotemark[3]Department of Mathematics and Geosciences,
    University of Trieste, Trieste, Italy.
}

@@

A tesseroid is a spherical mass element defined in geocentric spherical
coordinates
(Figure~\ref{fig:tesseroid}).
It is bounded by two meridians, two parallels, and two concentric circles.
The gravitational fields of a tesseroid at a point $P = (r,\phi,\lambda)$
are determined with respect to the local North-oriented coordinate system at
P (x, y, z in Figure~\ref{fig:tesseroid}).
\citet{Grombein2013} formulated Cartesian kernels for the volume integrals
that define the tesseroid gravitational potential, gravitational attraction,
and Marussi tensor, respectively,

@@

\noindent
where $\alpha, \beta \in \{x, y, z\}$,
$\rho$ is the density,
$G = 0.00000000006673\ m^3kg^{-1}s^{-1}$ is the gravitational constant,
$\kappa = {r^\prime}^2 \cos \phi^\prime$,
and

\begin{eqnarray}
    \Delta_x &=& r^\prime\cos\phi\sin\phi^\prime - \sin\phi\cos\phi^\prime
                 \cos(\lambda^\prime - \lambda), \\
    \Delta_y &=& r^\prime \cos \phi^\prime \sin(\lambda^\prime - \lambda), \\
    \Delta_z &=& r^\prime \cos \psi - r,\\
    \ell &=& \sqrt{{r^\prime}^2 + r^2 - 2 r^\prime r \cos \psi}, \\
    \cos\psi &=& \sin\phi\sin\phi^\prime + \cos\phi\cos\phi^\prime
                 \cos(\lambda^\prime - \lambda).
\end{eqnarray}

@@

The Gauss-Legendre Quadrature for three-dimensional volume integrals,
like equations
\ref{eq:tesspot}, \ref{eq:tessgrav}, and \ref{eq:tesstensor},
becomes \citep{Asgharzadeh2007}

@@

Applying equation~\ref{eq:glq3d} to equations
\ref{eq:tesspot}, \ref{eq:tessgrav}, and \ref{eq:tesstensor},
we see that $f(r_i, \phi_j, \lambda_k)$ is the effect of a point
mass located on the quadrature nodes.
Thus, it can be said that the GLQ integration
approximates the volume integrals  by a
weighted sum of point mass effects.

The accuracy of the integration
depends on the number of point masses used in the summation.
\citet{Ku1977} showed that it also depends on the ratio between
the distance to the computation point and the distance between adjacent nodes.
Figure~\ref{fig:glqerrorsample}
illustrates this effect on the $g_{xy}$ gravity gradient component.
The $g_{xy}$ component was produced by a
$7^\circ \times 7^\circ \times 20\ km$ tesseroid
with $2.67\ g.cm^{-3}$ density
and top at $z=0\ km$.
The maps were calculated on a regular grid
with $100\times100$ points.
Figure~\ref{fig:glqerrorsample}a shows the $g_{xy}$ component
calculated at 400 km height using
GLQ with order two (two point masses in the longitude, latitude, and vertical
directions, respectively).
Figure~\ref{fig:glqerrorsample}b shows $g_{xy}$ computed with order two
GLQ as well but at 150 km height.
Notice that the computed effect is concentrated around each point mass
of the GLQ (black dots) and does not resemble the effect of a tesseroid.
\citet{Ku1977} determined an empirical criterion that the distance between
point masses (quadrature nodes) should be smaller than the minimum distance to
the computation point.
Thus, if a computation point is too close to the tesseroid one would have to
decrease the distance between the point masses in order to obtain an accurate
result.
One way to accomplish this would be increase the order of the quadrature
$N$ in all three directions.
Figure~\ref{fig:glqerrorsample}c shows the $g_{xy}$ component calculated at
150km height but with a GLQ order of 30.
The computed $g_{xy}$ component more closely resembles the results from
\citet{Asgharzadeh2007}.

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/vary-height-and-order}
    \caption{
        Example of the effect of varying
        the computation height
        and the number of point masses in the Gauss-Legendre Quadrature
        (i.e., the GLQ order).
        Black circles represent the horizontal location of the GLQ nodes
        (not shown for GLQ order 30).
        a) Field calculated at $400\ km$ height using GLQ order 2.
        The effect is similar to that presented by \citet{Asgharzadeh2007}.
        b) At $150\ km$ height and order 2,
        the field resembles that of
        four point sources located at the GLQ nodes.
        This effect was shown by \citet{Ku1977}
        for the case of the $g_z$ component of a right rectangular prism.
        c) At $150\ km$ but with a higher GLQ order of 30,
        the field is similar to that expected for a single mass source.
    }
    \label{fig:glqerrorsample}
\end{figure}

@@

Instead of increasing the GLQ order,
\citet{Li2011} proposed an alternative method
for decreasing the distance between point masses
and achieve an accurate integration.

Instead of increasing the GLQ order,
they divide the tesseroid into smaller tesseroids while keeping the number of
point masses per tesseroid fixed.

The gravitational effect of the larger tesseroid would be calculated as the sum
of the effects of the eight smaller tesseroids.

This division would effectively decrease the distance between nodes because of
the smaller tesseroids.

The criterion for dividing a tesseroid is that
the distance to the computation point must be smaller than
a constant times the size of the tesseroid.

The size of the tesseroid is used as a proxy for the distance between point
masses.

This procedure is repeated recursively
(i.e., to each of the smaller tesseroids)
until all tesseroids are within the acceptable distance-size ratio
or a minimum size is achieved.

The advantage of this adaptive discretization is that the number of point
masses is only increased in regions of the tesseroid that are closer to the
computation point.

Alternatively, simply increasing the order of the GLQ would increase the number
of point masses evenly throughout the whole tesseroid.

@@

The calculation of the tesseroid gravitational fields with adaptive
discretization is implemented in the open-source package \textit{Tesseroids}.

Tesseroids is freely available under the BSD 3-clause free-software license.

An archived version of the source code is available along with this article.

The official website\footnote{http://www.leouieda.com/tesseroids}
hosts the software documentation and
archives of the latest binary executables and source code.

Ongoing development takes place through the source code hosting site
Github\footnote{https://github.com/leouieda/tesseroids}.

\textit{Tesseroids} is written in the C programming language.

Tesseroids consists of command-line programs.

Tesseroids tries to follow the Unix principals of
(1) making each program do one thing well,
(2) using plain text for input and output,
(3) making each program a ``filter'' that can be used in conjunction with other
programs.

The gravitational potential, attraction, and Marussi tensor components
are each computed by a separate program.

Programs for model and computation grid generation are also included.

The package also includes programs for computing the gravitational fields of
right rectangular prisms in Cartesian and spherical coordinates.

Tesseroids programs receive input through command-line arguments and the
standard input channel (STDIN) and output the results through
the standard output channel (STDOUT).

The command-line interface of each program was designed so that they
can be called in sequence using a Unix pipe.

For example, the command to
generate a regular grid with $NLON \times NLAT$ points,
calculate on it the $g_z$ and $g_{zz}$ caused by the tesseroids in file
MODELFILE,
and save the results to a file called OUTPUT
is:

@@

The \texttt{src} folder of the distribution archive
contains the C files that build the command-line programs
(e.g., \texttt{tessgz.c}).

The \texttt{src/lib} folder contains the source files that implement the
numerical computations.

We will not describe here the implementation of the input/output parsing and
other miscellanea.

Instead, we will focus on the details of the Gauss-Legendre Quadrature
integration of equations~\ref{eq:tesspot},~\ref{eq:tessgrav},
and~\ref{eq:tesstensor}
and the adaptive discretization of tesseroids.

All figures and numerical experiments in this article were produced by the
IPython notebook\footnote{http://ipython.org}
files \citep{Perez2007} in the supplementary material.

These files combine the analysis source code
with explanatory text, equations, and the figures generated by the code.

This supplementary material can be run and executed locally or viewed
online\footnote{https://github.com/pinga-lab/paper-tesseroids}
(in a non-executable form).

@@

The source file \texttt{src/lib/glq.c} contains the code necessary to perform a
Gauss-Legendre Quadrature integration.

The first step in the GLQ is to compute the
locations of the point masses (i.e., the discretization points).

These points are roots of Legendre polynomials.

Precomputed values for these roots are available low order polynomials,
usually up to order 5.

For flexibility, we use the multiple root-finder algorithm of
\citet{Barrera-Figueroa2006} to compute the roots for polynomials of any order.

The additional computational load is minimal since the root-finder algorithm
must be run only once per program execution.

The root-finder algorithm is implemented in functions \texttt{glq\_nodes} and
\texttt{glq\_next\_root}.

The computed roots will be in the range $[-1, 1]$
and must be scaled to the integration limits (the limits of the tesseroid)
using function \texttt{glq\_set\_limits} (see equation~\ref{eq:glq_scaling}).

The GLQ weights are computed using equation~\ref{eq:glq_weights} by function
\texttt{glq\_weights}.

Both the computed roots and weights are stored in a data structure (a C
\texttt{struct}) called \texttt{GLQ}.

Function \texttt{glq\_new} allocates the memory necessary,
calculates the roots and weights,
and returns the complete \texttt{GLQ} structure.

The numerical integration is performed by the functions in module
\texttt{src/lib/grav\_tess.c}.

The functions \texttt{tess\_pot}, \texttt{tess\_gx}, \texttt{tess\_gy},
and so on, compute the gravitational fields of a single tesseroid on a single
computation point using the GLQ.

These functions require three \texttt{GLQ} structures that contain the roots
and weights for GLQ integration in the three dimensions.

The roots must be scaled to the
integration limits $[\lambda_1, \lambda_2], [\phi_1, \phi_2], [r_1, r_2]$
(see equations~\ref{eq:tesspot}-\ref{eq:tesstensor}).

The integration consists of three \texttt{for} loops that sum the weighted
kernel functions evaluated at each GLQ point mass (the scaled roots).

The evaluation of the trigonometric functions in
equations~\ref{eq:tesspot}-\ref{eq:tesstensor} inside the inner loops
are the biggest bottlenecks for the numerical integration.

Better performance is achieved by
precomputing the sine and cossine of latitudes
and moving some trigonometric function
evaluations to the outer loops.

@@

The adaptive discretization algorithm of \citet{Li2011}
divides a tesseroid recursively.

A tesseroid will be divided when the smallest distance between it and the
computation point is smaller than a constant times the largest dimension of the
tesseroid.

Instead of the smallest distance,
we use in our implementation
the distance $d$ between
the computation point $(r, \lambda, \phi)$
and the geometric center of the tesseroid
$(r_t, \lambda_t, \phi_t)$

@@

We define the dimensions of the tesseroid
along longitude, latitude, and radius, respectively, as
(see Figure~\ref{fig:division}a)

@@

We fix the GLQ order $N=2$ for all three dimensions, resulting in eight point
masses in total.

The criterion we use to determine if a tesseroid must be divided is

\begin{equation}
    \frac{d}{L_i} < D,
    \label{eq:condition}
\end{equation}
\noindent
where $i \in (\lambda, \phi, r)$
and $D$ is a positive scalar called the ``distance-size ratio''.

$D$ determines how close the
computation point can be before we must divide the tesseroid.

Thus, the value of $D$ is indirectly responsible for
the accuracy of the solution and the computation time.

The modeling programs provide default values of $D$.

It can also be controlled through command-line arguments passed to the
programs.

We used a stack-based implementation of the algorithm instead of using
recursive function calls, as originally proposed by \citet{Li2011}.

Stacks are array-like data structures with a particular way of inserting and
removing elements from it.

In a stack, one can only insert elements on the top (the last empty position).

Likewise,
one can only remove the last element that was placed on the stack
(commonly referred to as ``popping'' the stack).

Stacks are also known as ``Last-in-first-out'', or LIFO, structures.

LIFOs are often described as being analogous to a stack of plates.

One can only place a new plate on the top of the stack and
can only remove a plate safely from the top of the stack.

We will describe the stack-based algorithm for a single tesseroid and
computation point.

The algorithm is implemented in function \texttt{calc\_tess\_model\_adapt}
of the file \texttt{src/lib/grav\_tess.c}.

Start by creating an empty stack of tesseroids.

The  stack is represented by the \texttt{stack} variable,
an array of \texttt{TESSEROID} structures.

We must define a maximum size for the stack to allocate memory for it.

Defining a maximum size allows us to avoid an infinite loop and large
consumption of computer memory in case the computation point is
on (or sufficiently close to) the surface of the tesseroid.

We use the integer \texttt{stktop} to keep track of the index of the last
element in the stack (the top of the stack).

The stack is initialized with the input tesseroid.

The initialization is done by copying the tesseroid into the stack and
setting \texttt{stktop} to zero (the first element).

Once the stack is initialized, the steps of the algorithm are:

\begin{enumerate}
    \item ``Pop'' the stack (i.e., take the last element from it).
        This will cause \texttt{stktop} to be reduced by one.
    \item Compute the distance $d$ between the tesseroid and
        the computation point (equation~\ref{eq:distance}).
    \item Compute the dimensions of the tesseroid $L_\lambda$, $L_\phi$,
        and $L_r$ (equations~\ref{eq:sizelon}-\ref{eq:sizer}).
    \item Check the condition in equation~\ref{eq:condition} for each
        dimension of the tesseroid.
    \item If none of the dimensions fail the above condition,
        compute the gravitational effect of the tesseroid using the
        Gauss-Legendre Quadrature with order two
        (equation~\ref{eq:glq3d}).
    \item If any of the dimensions fail the condition:
    \begin{enumerate}
        \item Divide the tesseroid in half along each dimension that failed
             the condition.
        \item Check if the number of new elements plus \texttt{stktop}
             is smaller than the maximum stack size.
             If it is not (i.e., the stack is full),
             warn the user of a ``stack overflow''
             and compute the effect of the larger tesseroid (step 5).
        \item If the stack is not full, place the smaller tesseroids into
             the stack.
    \end{enumerate}
    \item Repeat the above steps until the stack is empty
        (\texttt{stktop} is equal to -1).
\end{enumerate}

This stack-based implementation has some advantages over the original recursive
implementation suggested by \citet{Li2011}.

(1) It will be faster because it bypasses the overhead of function calls.

The speedup will be specially significant if the implementation is done
in high level interpreted languages, like Matlab and Python.

(2) It gives the developer more control over the recursion step.

In recursive implementations,
the developer has no control over the maximum number of consecutive
recursive calls (the ``recursion depth'').

This limit may vary with programming language, compiler, and operating system.

Overflowing the maximum recursion depth usually results in program
crashes with cryptic or inexistent error messages.

In the stack-based implementation, the developer has complete control.

Overflowing of the stack can be handled gracefully with an error message or
even performing a suitable approximation of the result.

Figure~\ref{fig:division} shows the effect of the distance-size ration $D$
in the adaptive discretization of a tesseroid for the computation at a point P.

Computing the gravitational effect of the tesseroid in
Figure~\ref{fig:division}a using the algorithm described above
is equivalent to computing the effect of the tesseroids in
Figures~\ref{fig:division}b-d.

Note that increasing the value of $D$ results in a finer mesh of tesseroids.

For $D = 1$ (Figure~\ref{fig:division}b) the number of tesseroids in the
adaptive discretization is 4,
for $D = 2$ (Figure~\ref{fig:division}c) it is 38,
and for $D = 6$ (Figure~\ref{fig:division}d) it is 936.

@@

>>>>

\author{
    Leonardo Uieda\footnotemark[1]\footnotemark[2],
    et al.
}

\address{
    \footnotemark[1]Universidade do Estado do Rio de Janeiro,
    Rio de Janeiro, Brazil.
    email: leouieda@gmail.com
}

@@

A tesseroid is a mass element defined in geocentric spherical
coordinates
(Figure~\ref{fig:tesseroid}).
It is bounded by two meridians, two parallels, and two concentric circles.
The gravitational fields of a tesseroid at a point $P = (r,\phi,\lambda)$
are determined with respect to the local North-oriented coordinate system at
P (x, y, z in Figure~\ref{fig:tesseroid}).
\citet{Grombein2013} formulated Cartesian kernels for the volume integrals
that define the tesseroid gravitational potential, gravitational attraction,
and Marussi tensor, respectively,

@@

\noindent
where $\alpha, \beta \in \{x, y, z\}$,
$\rho$ is the density,
$G = 6.673\times10^{-11}\ m^3kg^{-1}s^{-1}$ is the gravitational constant,
and

\begin{eqnarray}
    \Delta_x &=& r^\prime\cos\phi\sin\phi^\prime - \sin\phi\cos\phi^\prime
                 \cos(\lambda^\prime - \lambda), \\
    \Delta_y &=& r^\prime \cos \phi^\prime \sin(\lambda^\prime - \lambda), \\
    \Delta_z &=& r^\prime \cos \psi - r,\\
    \kappa &=& {r^\prime}^2 \cos \phi^\prime, \\
    \ell &=& \sqrt{{r^\prime}^2 + r^2 - 2 r^\prime r \cos \psi}, \\
    \cos\psi &=& \sin\phi\sin\phi^\prime + \cos\phi\cos\phi^\prime
                 \cos(\lambda^\prime - \lambda).
\end{eqnarray}

@@

The Gauss-Legendre Quadrature for three-dimensional volume integrals,
like equations~\ref{eq:tesspot}-\ref{eq:tesstensor},
becomes \citep{Asgharzadeh2007}

@@

Comparing equation~\ref{eq:glq3d} with
equations~\ref{eq:tesspot}-\ref{eq:tesstensor},
we see that $f(r_i, \phi_j, \lambda_k)$ is the effect of a point
mass located on the quadrature nodes.
Thus, it can be said that the GLQ integration
approximates the volume integrals  by a
weighted sum of point mass effects.

The accuracy of the integration
depends on the number of point masses used in the summation.
\citet{Ku1977} showed that it also depends on the ratio between
the distance to the computation point and the distance between adjacent nodes.
Figure~\ref{fig:glqerrorsample}
illustrates this effect on the $g_{xy}$ gravity gradient component.
The $g_{xy}$ component was produced by a
$7^\circ \times 7^\circ \times 20\ km$ tesseroid
with $2.67\ g.cm^{-3}$ density
and top at $z=0\ km$.
The maps were calculated on a regular grid
with $100\times100$ points.
Figure~\ref{fig:glqerrorsample}a shows the $g_{xy}$ component
calculated at 400 km height using
GLQ with order two ($2 \times 2 \times 2 = 8$ point masses).
Figure~\ref{fig:glqerrorsample}b shows $g_{xy}$ computed with order two
GLQ as well but at 150 km height.
Notice that the computed effect is concentrated around each point mass
of the GLQ (black dots) and does not resemble the effect of a tesseroid.
\citet{Ku1977} determined an empirical criterion that the distance between
point masses (quadrature nodes) should be smaller than the minimum distance to
the computation point.
Thus, if a computation point is too close to the tesseroid one would have to
decrease the distance between the point masses in order to obtain an accurate
result.
One way to accomplish this would be increase the order of the quadrature
$N$ in all three directions.
Figure~\ref{fig:glqerrorsample}c shows the $g_{xy}$ component calculated at
150km height but with a GLQ order of 30
($30 \times 30 \times 30 = 27,000$ point masses).
The computed $g_{xy}$ component more closely resembles
the expect results for a single tesseroid \citep{Asgharzadeh2007}.

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figs/vary-height-and-order}
    \caption{
        Example of the effect of varying
        the computation height
        and the number of point masses in the Gauss-Legendre Quadrature.
        Black circles represent the horizontal location of the point masses.
        a) $g_{xy}$ calculated at $400\ km$ height using GLQ order 2
        ($2 \times 2 \times 2 = 8$ point masses).
        b) At $150\ km$ height and GLQ order 2,
        the result resembles that of
        four point mases instead of a single tesseroid.
        This effect was shown by \citet{Ku1977}.
        c) At $150\ km$ but with a higher GLQ order of 30,
        the result is similar to that expected for a single mass source.
    }
    \label{fig:glqerrorsample}
\end{figure}

@@

\citet{Li2011} proposed an alternative method
for decreasing the distance between point masses
and achieve an accurate integration.
Instead of increasing the GLQ order,
they keep it fixed to a given number
and divide the tesseroid into smaller parts.
The sum of the effects of the smaller tesseroids
is equal to the gravitational effect of the larger tesseroid.
This division effectively decreases
the distance between nodes because
of the smaller size of the tesseroids.
The criterion for dividing a tesseroid or not
is that the distance to the computation point
must be smaller than a constant times the size of the tesseroid.
This is analogous to the criterion of \citet{Ku1977}
because the size of the tesseroid serves as a proxy
for the distance between point masses.
This procedure is repeated recursively
(i.e., to each of the smaller tesseroids and so on)
until all tesseroids are within the acceptable
ratio of distance and size or a minimum size is achieved.

The advantage of this adaptive discretization is
that the number of point masses is only increased
in regions of the tesseroid that are
closer to the computation point.
Alternatively,
simply increasing the order of the GLQ
would increase the number of point masses
evenly throughout the whole tesseroid.

@@

We have implemented the calculation of
the tesseroid gravitational fields
with adaptive discretization
in the open-source package \textit{Tesseroids}.
The software is freely available
on the Internet\footnote{http://www.leouieda.com/tesseroids}
under the BSD 3-clause license.
An archived version of the source code
is available as part of this article.

\textit{Tesseroids}  consists of command-line programs
written in the C programming language.
The command-line interface follows the Unix principals of
(1) making each program do one thing well,
(2) using plain text for input and output,
(3) making each program a ``filter'' that can be used in conjunction with other
programs.
The package includes programs to calculate
the gravitational fields of tesseroids and
rectangular prisms (in both Cartesian and spherical coordinates),
as well as model and computation grid generation.
All programs receive input through
command-line arguments and the standard input channel (STDIN)
and output the results through the standard output channel (STDOUT).
For example,
the command to generate a regular grid with $NLON \times NLAT$ points,
calculate on it $g_z$ and $g_{zz}$ caused by
the tesseroids in a file ``MODELFILE'',
and save the results to a file called ``OUTPUT''
is:

@@

The \texttt{src} folder of the source code archive
contains the C files that build the command-line programs
(e.g., \texttt{tessgz.c}).
The \texttt{src/lib} folder contains
the source files that implement the numerical computations.
We will not describe here the implementation of the input/output parsing and
other miscellanea.
Instead, we will focus on the details of the Gauss-Legendre Quadrature
integration of equations~\ref{eq:tesspot}-\ref{eq:tesstensor}
and the adaptive discretization of tesseroids.

All figures and numerical experiments in this article
were produced by IPython notebooks\footnote{http://ipython.org}
\citep{Perez2007}.
These files combine the analysis and figure generation source code
with text, equations, and the figures generated by the code.
These files,
as well as instructions for installing and running the analysis,
are distributed along with the source code archive
that accompanies this article.
Alternatively,
the notebooks and can also be downloaded
or viewed in a non-executable form from
an online repository\footnote{https://github.com/pinga-lab/paper-tesseroids}.

@@

The source file \texttt{src/lib/glq.c}
contains the code necessary to perform
a Gauss-Legendre Quadrature integration.
The first step in the GLQ is to compute the
locations of the discretization points (i.e., the point masses).
These points are roots of Legendre polynomials.
Precomputed values are available low order polynomials,
typically up to order five.
For flexibility and to compute higher order roots,
we use the multiple root-finder algorithm of
\citet{Barrera-Figueroa2006}.
The additional computational load is minimal
because the root-finder algorithm
must be run only once per program execution.
The root-finder is implemented in functions
\texttt{glq\_nodes} and \texttt{glq\_next\_root}.
The computed roots will be in the range $[-1, 1]$
and must be scaled to the integration limits
(the physical boundaries of the tesseroid)
using function \texttt{glq\_set\_limits} (see equation~\ref{eq:glq_scaling}).

The GLQ weights (equation~\ref{eq:glq_weights})
are computed by function \texttt{glq\_weights}.
Both the computed roots and weights are stored in a data structure
(a C \texttt{struct}) called \texttt{GLQ}.
Function \texttt{glq\_new}
handles memory allocation,
calculates the roots and weights,
and returns the complete \texttt{GLQ} structure.

The numerical integration of the tesseroid gravitational fields
is performed by the functions in module \texttt{src/lib/grav\_tess.c}.
Functions \texttt{tess\_pot}, \texttt{tess\_gx}, \texttt{tess\_gy}, and so on,
compute the gravitational fields of a single tesseroid
on a single computation point.
These functions require three \texttt{GLQ} structures,
each containing the roots and weights
for GLQ integration in the three dimensions.
The roots must be scaled to the
integration limits
$[\lambda_1, \lambda_2], [\phi_1, \phi_2], [r_1, r_2]$
(see equations~\ref{eq:tesspot}-\ref{eq:tesstensor}).
The integration consists of three loops
that sum the weighted kernel functions
evaluated at each GLQ point mass (the scaled roots).

The biggest bottlenecks for the numerical integration are
the number of point masses used
and he evaluation of the trigonometric functions in
equations~\ref{eq:tesspot}-\ref{eq:tesstensor} inside the inner loops.
Better performance is achieved
by pre-computing the sine and cosine of latitudes
and moving some trigonometric function evaluations
to the outer loops.

@@

Our implementation of the adaptive discretization algorithm
differs in a few ways from the original.
In \citet{Li2011},
a tesseroid will be divided when
the smallest distance between it and the computation point
is smaller than a constant times
the largest dimension of the tesseroid.
Instead of the smallest distance,
we use the easier to calculate
distance between
the computation point $(r, \lambda, \phi)$
and the geometric center of the tesseroid
$(r_t, \lambda_t, \phi_t)$

@@

Our definition of the dimensions of the tesseroid
(the ``side lengths'' of \citet{Li2011})
along longitude, latitude, and radius, respectively, are
(Figure~\ref{fig:division}a)

@@

To determine if a tesseroid must be divided,
we check if

\begin{equation}
    \frac{d}{L_i} < D,
    \label{eq:condition}
\end{equation}

\noindent
for each $i \in (\lambda, \phi, r)$.
$D$ is a positive scalar
hereafter referred to as the ``distance-size ratio''.
If the inequality holds for all three dimensions,
the tesseroid is not divided.
Thus, the distance-size ratio determines
how close the computation point can be
before we must divide the tesseroid.
The value of $D$ is indirectly responsible for
the accuracy of the solution and the computation time.
We will explore the relationship with the accuracy in the following section.

Figure~\ref{fig:division} shows examples of
the resulting tesseroid models after adaptive discretization.
Figure~\ref{fig:division}a shows
the initial tesseroid and computation point P.
Figures~\ref{fig:division}b-d are
the result of adaptive discretization using
different values of the distance-size ratio $D$,
respectively,
$D=1$, $D=2$, and $D=6$.
The number of tesseroids in the resulting discretization is, respectively,
4, 38, and 936.
Code to perform the adaptive discretization
and produce Figure~\ref{fig:division} is available in
the IPython notebook \texttt{methods\_figures.ipynb}.

@@

Instead of using recursive function calls,
as originally proposed by \citet{Li2011},
we use a stack-based implementation of the algorithm.
Stacks are array-like data structures
with a particular way of inserting and removing elements from it.
In a stack,
one can only insert elements to the top of the stack
(the last empty position).
Likewise,
one can only remove the last element of the stack
(commonly referred to as ``popping'' the stack).
Stacks are also known as ``Last-in-first-out'', or LIFO, structures.

We will describe the stack-based algorithm
for computing the effect of a single tesseroid
on a single computation point.
The algorithm is implemented in function \texttt{calc\_tess\_model\_adapt}
of the file \texttt{src/lib/grav\_tess.c}.
The stack of tesseroids is represented by
the \texttt{stack} variable,
an array of \texttt{TESSEROID} structures.
We must define a maximum size for the stack to allocate memory for it.
Defining a maximum size allows us to
avoid an infinite loop
in case the computation point is on
(or sufficiently close to) the surface of the tesseroid.
We use the integer \texttt{stktop}
to keep track of the index of
the last element in the stack (the top of the stack).

The algorithm starts by creating an empty stack of tesseroids.
Then, the stack is initialized with the input tesseroid.
The initialization is done by copying the tesseroid into the stack
and setting \texttt{stktop} to zero (the first element).
Once the stack is initialized, the steps of the algorithm are:

\begin{enumerate}
    \item ``Pop'' the stack (i.e., take the last tesseroid from it).
        This will cause \texttt{stktop} to be reduced by one.
        This tesseroid is the one that will be evaluated in the following
        steps.
    \item Compute the distance $d$ between
        the geometric center of the tesseroid and
        the computation point using equation~\ref{eq:distance}.
    \item Compute the dimensions of the tesseroid $L_\lambda$, $L_\phi$,
        and $L_r$ using equations~\ref{eq:sizelon}-\ref{eq:sizer}.
    \item Check the condition in equation~\ref{eq:condition} for each
        dimension of the tesseroid.
    \item If none of the dimensions fail the above condition,
        compute the gravitational effect of the tesseroid using the
        Gauss-Legendre Quadrature
        (equations~\ref{eq:tesspot}-\ref{eq:tesstensor} and~\ref{eq:glq3d}).
        We use a GLQ order of two for all three dimensions
        ($2 \times 2 \times 2 = 8$ point masses)
        by default.
        This value can be changed using
        a command-line argument of the modeling programs.
    \item If any of the dimensions fail the condition:
    \begin{enumerate}
        \item Divide the tesseroid in half along each dimension that failed
             the condition.
        \item Check if the number of new elements plus \texttt{stktop}
             is smaller than the maximum stack size.
             If it is not (i.e., the stack is full),
             warn the user of a ``stack overflow''
             and compute the effect of the larger tesseroid (step 5).
        \item If the stack is not full, place the smaller tesseroids into
             the stack.
    \end{enumerate}
    \item Repeat the above steps until the stack is empty
        (\texttt{stktop} is equal to -1).
\end{enumerate}

This stack-based implementation
has some advantages over the original recursive implementation,
namely:
(1) It will be faster because it bypasses the overhead of function calls.
(2) It gives the developer more control over the recursion step.
In recursive implementations,
the developer has no control over
the maximum number of consecutive recursive calls
(i.e., the ``recursion depth'').
This limit may vary with programming language,
compiler, and operating system.
Overflowing the maximum recursion depth
may result in program crashes,
typically with cryptic or inexistent error messages.
In the stack-based implementation,
the developer has complete control.
Overflowing of the stack can be handled gracefully
with an error message
or even performing a suitable approximation of the result.
